---
title: "Introduction to R"
author: "Task"
---

1. Load the `HappinessAlcoholConsumption.csv` dataset that you can find in the `S2/data/` folder.

```{r}
db <- read.csv("data/HappinessAlcoholConsumption.csv")
```


2. Look at the structure of the dataset

```{r}
str(db)
```
So, here we can see that we have four numeric variable, three characters one (that we must convert into factors) and one integer.

Let's convert the character variables as factors:

```{r}
db$Region <- as.factor(db$Region)
db$Country <- as.factor(db$Country)
db$Hemisphere <- as.factor(db$Hemisphere)
```

Another useful function is `summary()`:

```{r}
summary(db)
```
Here, we can see some position indices for the numerical variables, and the absolute frequency for each level for the factor (categorical) variables.


3. How many observations? How many variables?

There are many ways to understand the number of observations and variables:

```{r}
dim(db) #dimension of the dataframe (rows: observations, columns: variables)
```
```{r}
nrow(db) #number of rows (i.e., observations)
```


```{r}
ncol(db) #number of columns (i.e., variables)
```


4. Create another dataset containing the variables `Country` and `HappinessScore`

```{r}
db_sel <- db[,c("Country", "HappinessScore")]
```


5. Compute the mean of the `HappinessScore` score for each region

One solution:

```{r}
lev <- levels(db$Region)

mean(db$HappinessScore[db$Region == lev[1]])
mean(db$HappinessScore[db$Region == lev[2]])
mean(db$HappinessScore[db$Region == lev[3]])
mean(db$HappinessScore[db$Region == lev[4]])
mean(db$HappinessScore[db$Region == lev[5]])
mean(db$HappinessScore[db$Region == lev[6]])
mean(db$HappinessScore[db$Region == lev[7]])
mean(db$HappinessScore[db$Region == lev[8]])
mean(db$HappinessScore[db$Region == lev[9]])


```


another one:

```{r}
lev <- levels(db$Region)

for(i in seq(length(lev))){
  
  mean(db$HappinessScore[db$Region == lev[i]])
}

```

another one:

```{r}
library(tidyverse)

db %>%
  group_by(Region) %>%
  summarise(mean_region = mean(HappinessScore))
```



6. How many country has a mean below the global mean?

```{r}
global_mean <- mean(db$HappinessScore)

sum(db$HappinessScore< global_mean)
```


7. Create a new variable as the sum of `BeerPerCapita`, `SpiritPerCapita` and `WinePerCapita`


```{r}
db$new_var <- db$Beer_PerCapita + db$Spirit_PerCapita + db$Wine_PerCapita
```



8. Compute the median of this new variable considering only the north hemisphere.


One solution:

```{r}
median(db$new_var[db$Hemisphere == "north"])
```
another one:

```{r}
db %>%
  filter(Hemisphere == "north") %>%
  summarise(median_new_var = median(new_var))
```

